---
title: "Unemployment and Level of Education"
author: |
  Melle Kooge 2855016
  Lex Veerman 2821630
  Michael Minneboo 2858823
  Mateo van Gerven 2826991
  Merel Vonk 2862870
  Kirill Müller 2831153
  Athiraiyan Visvalingam 2801095

output: pdf_document
date: "2025-06-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r load_data}
source("controle_data.R") 
```

# Title Page

# Part 1
Identify a Social Problem The social problem we have chosen is that low-educated people are more often unemployed than high-educated people. In this report we will only focus on statistics in the Netherlands.

## 1.1 Motivation
The inequality between lower and higher educated people is a continuing challenge in the labor market. According to CBS (Centraal Bureau voor de Statistiek, a surveyor of economic data from mainly the Netherlands), lower educated people are more frequently unemployed than higher educated peers. This will cause a broadening gap between lower and higher educated employees, impacting both their opportunities in the labor market and their financial stability. This research is aimed to store our findings in R to give a general overview of the social problem.

# Part 2 - Data Sourcing

## 2.1 Load in the Data

```{r loading_data, include=FALSE}
dataset_1 <- Educationlevel2015
url ("https://opendata.cbs.nl/#/CBS/nl/dataset/85525NED/table")
dataset_2 <- Unemployment
url ("https://www.cbs.nl/nl-nl/visualisaties/dashboard-arbeidsmarkt/werklozen/werkloosheid-per-provincie")
```

## 2.2 Summary of the data

```{r data_summary}
head(dataset_1)
#In this data set the different level of education has been measured per Region (provinces) per year in the Netherlands. It is a percentage based on the total inhabitants per region.

head(dataset_2)
#In this dataset the unemployment has been measured per Region (provinces) per year in the Netherlands. It is a percentage based on the working population. Currently the data is measured per quarter of a year. We need to have an average from each year to have it corresponds with our other dataset. The years that have been measured also differs with the other dataset, so we only use the years that both datasets inlcude.
```

## 2.3 Describe the variables included

```{r}
head(dataset_1)
#The variables we have in this data set are: 
  ##level of education:the different levels of education people have in the Netherlands 
  ##region: the twelve provinces in the Netherlands
  ##year: the years go through 2013 to 2023
head(dataset_2)
#The variables we have in this data set are:
  ##level of unemployment: the percentage of people that are unemployed from the working population
  ##region: the twelve provinces in the Netherlands
  ##quarter per year: the years go through the first quarter of 2015 to the first quarter of 2025

```

#Part 3 Quantifying

## 3.1 Data cleaning

```{r}
#Load the dataset
Unemployment <- read.csv2("Werkloosheidpercentage.csv") 

#Rename the first column and extract the year from the 'Quarter' column 
colnames(Unemployment)[1] <- "Quarter" 
Unemployment$Year <- str_extract(Unemployment$Quarter, "\\d{4}") 

#Replace commas with dots in the numeric columns and convert to numeric 
#(exclude 'Quarter' and 'Jaar' columns) 
numeric_cols <- setdiff(colnames(Unemployment), c("Quarter", "Year")) 
Unemployment[numeric_cols] <- lapply(Unemployment[numeric_cols], function(x) { 
  as.numeric(str_replace_all(x, ",", ".")) 
}) 

#Group by year and calculate the mean per column 
Unemployment_yearly <- Unemployment %>% 
  group_by(Year) %>% 
  summarise(across(all_of(numeric_cols), ~round(mean(.x, na.rm = TRUE), 2))) 

#Optional: Export to CSV 
write.csv(Unemployment_yearly, "Unemployment_by_year.csv", row.names = FALSE)

```

The file is created where the total unemployment per year with columns is clear. Firstly the first column was renamed to Quarter. Then the years from the data set were extracted, so they could be grouped by year. Previously the data was grouped per quarter per year, so to get a yearly average the mean from each year per region was calculated. To calculate the mean, the necessary numbers were converted to numeric.

```{r}
Unemployment_yearly <- Unemployment_yearly[-c(9, 10, 11), ]
```

The rows for the years 2023, 2024 and 2025 were removed, as these years were not used to make the necessary computations.

```{r}
#Load in the education level by year datasets
Educationlevel2015=read.csv2("2015_onderwijsniveau.csv") 
Educationlevel2016=read.csv2("2016_onderwijsniveau.csv") 
Educationlevel2017=read.csv2("2017_onderwijsniveau.csv") 
Educationlevel2018=read.csv2("2018_onderwijsniveau.csv") 
Educationlevel2019=read.csv2("2019_onderwijsniveau.csv") 
Educationlevel2020=read.csv2("2020_onderwijsniveau.csv") 
Educationlevel2021=read.csv2("2021_onderwijsniveau.csv") 
Educationlevel2022=read.csv2("2022_onderwijsniveau.csv") 
Educationlevel2023=read.csv2("2023_onderwijsniveau.csv") 
Educationlevel_by_year=rbind(Educationlevel2015, Educationlevel2016, Educationlevel2017, Educationlevel2018, Educationlevel2019, Educationlevel2020, Educationlevel2021,Educationlevel2022,Educationlevel2023) 
write.csv(Educationlevel_by_year,"Total_educationlevel.csv", row.names = FALSE) 
```

The table with education levels were sorted by year. So, the tables were downloaded by year and then bounded.

```{r}
#Renamed all the column names to an easier accessible name or to an English word
Educationlevel_by_year <- Educationlevel_by_year %>%  
  rename( 
    Basisonderwijs = Onderwijsniveau.5.categorieën.11.Basisonderwijs...., 
    VMBO_HAVO_VWO_Onderbouw_MBO_1 = Onderwijsniveau.5.categorieën.12.Vmbo..havo...vwo.onderbouw..mbo1...., 
    HAVO_VWO_MBO_2_4 = Onderwijsniveau.5.categorieën.21.Havo..vwo..mbo2.4...., 
    HBO_WO_Bachelor = Onderwijsniveau.5.categorieën.31.Hbo...wo.bachelor...., 
    HBO_WO_Master_Doctor = Onderwijsniveau.5.categorieën.32.Hbo...wo.master..doctor....,
    Gender = Geslacht,
    Age = Leeftijd,
    Year = Perioden,
    Region = Regio.s,
    Population = Bevolking..aantal.
  )
```

The column names in the dataset for the education level by year were simplified and changed to English, to make improve the readability of the dataset.

```{r}
Low_and_High_Education <- Low_and_High_Education[, -c(6, 7, 8, 9, 10)] 
```

Some columns were also deleted to improve the readability. This piece of code was not the only instance when this was used.

## 3.2 Generate necessary variables

```{r}
Low_and_High_Education <- Educationlevel_by_year %>% 
  mutate( 
    Low_Education_rate = 
      Basisonderwijs + 
      VMBO_HAVO_VWO_Onderbouw_MBO_1 + 
      HAVO_VWO_MBO_2_4, 
    High_Education_rate =  
      HBO_WO_Bachelor + 
      HBO_WO_Master_Doctor 
  ) 
```

The used datasets divided the education level in 5 different parts. To make the computations easier two variables were created, dividing these 5 parts into low education and high education. This was done by adding the percentages in the different parts. Now there were only two variables to use in the computations.

```{r}
#Added the Unemployment_by_Low_Education column
Education_Unemployment_yearly_transposed_provinces <- Education_Unemployment_yearly_transposed_provinces %>% 
  mutate(Unemployment_by_Low_Education = Unemployment / Low_Education_rate) 
```

The third variable was created by dividing the unemployment by low education rate. This was done to realise a variable to find a connection between the unemployment and the low education rate in each province. By using unemployment and the low education rate, intuitively there should be a high unemployment by low education rate in provinces with a low education rate, and it should be low in provinces with a high education rate. 

## 3.3 Visualize temporal variation

```{r}
ggplot(Education_Employment, aes(x = Year, y = Unemployment_by_Low_Education)) + 
  geom_line() + 
  labs(x = "\nYear", y = "Unemployment \nper Low Educated\n")+ 
  scale_x_continuous(breaks = seq(2015, 2022, by = 1), lim = c(2015, 2022)) 
```

## 3.4 Visualize spatial variation

```{r}
ggplot(map_data, aes(fill = Unemployment_by_Low_Education)) + 
  geom_sf(color = "white", size = 0.2) + 
  scale_fill_gradient(low = "blue", high = "red", 
                      name = "Unemployed per\nLow Educated\nin 2022") 
```

## 3.5 Visualize sub-population variation

```{r}
ggplot(data = mean_by_provinces, mapping = aes(x = Region_Group, y = Mean_Unemployment_by_Low_Education)) + 
  geom_boxplot() + 
  labs(x = "\nRegion", y = "Unemployment by \nLow Educated\n")
```

## 3.6 Event analysis

```{r}
ggplot(Education_Employment, aes(x = Year, y = Unemployment_by_Low_Education)) + 
  geom_line() + 
  geom_vline(xintercept = 2020,  
             color = "red",  
             linetype = "dashed") + 
  labs(x = "\nYear", y = "Unemployment \nper Low Educated\n")+ 
  scale_x_continuous(breaks = seq(2015, 2022, by = 1), lim = c(2015, 2022))
```


# Part 4 Discussion

At the start of our research, we hypothesized that there would be a clear correlation between education level and unemployment rates within the Netherlands. Specifically, we expected that regions with a lower average education level would have higher unemployment, while regions with a higher education level would experience lower unemployment. This expectation was based on the general assumption that higher education provides individuals with better job opportunities and employability.
However, after processing the datasets and visualizing the results in graphs, our findings challenged this expectation. For example, the province of Zeeland had the highest proportion of low-educated individuals, yet it also had the lowest unemployment rate of all provinces. This was a surprising outcome that did not align with our initial hypothesis.
Another notable finding was the contrast between the Randstad region and the northern provinces of the Netherlands. Despite the Randstad being more urbanized and economically developed, the unemployment rate among low-educated individuals was higher in the Randstad than in the northern provinces. This was unexpected, as we initially assumed that urban areas with more diverse economies would offer more opportunities, even for those with lower levels of education.



# Part 5 Reproducibility

## 5.1 Github repository link
https://github.com/Lex0803/project.git 

## 5.2 Reference List
CBS Statline. (z.d.). https://opendata.cbs.nl/#/CBS/nl/dataset/85525NED/table

Centraal Bureau voor de Statistiek. (z.d.). Werkloosheid per provincie. Centraal Bureau Voor de Statistiek. https://www.cbs.nl/nl-nl/visualisaties/dashboard-arbeidsmarkt/werklozen/werkloosheid-per-provincie 

